---
title: "Analyse visa-free network data with multilevel (mixed, hierarchical) models"
author: "Rouslan Karimov"
date: "Last updated on `r format(Sys.time(), '%d %B %Y')`"
project: dissertation
output: html_notebook
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(root.dir = '/Users/rkarimov/Dropbox/GradSchool/Dissertation/data_derived/', echo = TRUE)
```


# Purpose

This is a notebook where I prepare data for further analysis.

## Data munging

```{r message=FALSE, warning=FALSE}
rm(list = ls())

# Load libraries
library(tidyverse)
library(here)

# Load master dataset
dta <- read_csv(here('data_derived', 'master_dataset_all_vars.csv'), 
                col_types = cols(max.stay.days = 'd', langdist = 'd',
                                 migr.stock = 'd', secval.diff = 'd',
                                 secval.dest = 'd',
                                 region.dest = 'f', region.orig = 'f',
                                 income.group.dest = 'f', orig.income.group = 'f',
                                 contig = 'l', orig.was.colony = 'l',
                                 dest.was.colony = 'l'))

head(dta)
```

For programming convenience, I am going to drop some variables as unnecessary. There include:

-   Geographic regions (collinear with distance, culture and GDP)
-   Income groups (collinear with GDP)
-   Origin-specific secular values scores. The hypothesis is that they don't affect a destination's openness. Their similarity to the destination will be captured by the difference in secular values, which I keep.
-   Origin GDP/capita. We have destination GDP and GDP difference; the origin GDP is redundant.
-   Total GDP or origin: not used in model.
-   Origin-specific democracy scores. We have destination demscores and the destination/origin difference; the origin demscore is redundant.
-   Destination-specific fragility scores. The hypothesis is that it's the origin's fragility that makes its citizens into likely migrants/refugees, and so will invite stricter visa regimes from destinations. Also, we keep fragility difference, so destination fragility is redundant.
-   Unadjusted number of travelers from OAG.

```{r}
dta_compact <- select(dta, -gdp.total.orig, -gdp.per.cap.orig, -region.dest, -region.orig, -income.group.dest, -income.group.orig, -secval.orig, -demscore.orig, -numtravelers.oag, -fragility.dest)
```

I am also going to scale and center everything that's not binary or categorical and isn't already scaled. This will make sure the model converges and that the constants have a nicer interpretation (effect when predictors are fixed at their means).

```{r}
scaled <- scale(select(dta_compact, distw, gdp.per.cap.dest, gdp.total.dest, gdp.per.cap.diff,  langdist,
                       secval.dest, secval.diff, fdi.stock, exports.from.orig, pop.thou.dest, migr.stock,
                       numtravelers.oag.adj))
scaleddta <- data.frame(select(dta_compact, -colnames(scaled)), scaled)
```

# Export data

```{r}
save(scaleddta, file = here('data_derived', 'scaled.RData'))
```



